---
title: "DEG_analysis_for_spine_samples"
output: html_document
---

necessary libraries are loaded.
```{r}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
library(DESeq2)
library(pheatmap)
```

working directory is set.
```{r}
setwd("/Users/onat/Desktop/project/spine/spine_count_data")
```

count data is loaded.
```{r}
data <- read.csv(file = "spine_count_data.csv", sep = ";", header=T,fill=T,row.names = NULL)
colnames(data) <- c(" ","SRR5714742","SRR5714765","SRR5714766","SRR5714770","SRR5714774","SRR5714740","SRR5714792","SRR5714793","SRR5714794","SRR5714796")
data2 <- data[,-1]
rownames(data2) <- make.names(data[,1], unique = T)
rm(data)
```

non zero gene matrix is created.
```{r}
no_zero_genes_mtx <- data2[!rowSums(data2==0), ]
```


NA values are converted to 0 and filter out lowly expressed transcripts** 
```{r}

data2[!rowSums(data2<100), ]
t(data2[!rowSums(data2<5000),][1,])
data2[is.na(data2)] <- 0
data2 <- data2[apply(data2, 1, function(x) length(x[x >= 5]) >= 5),]
```

meta data is loaded.
```{r}
tissue <- c("spine", "spine", "spine","spine", "spine","turbinate" ,"turbinate", "turbinate","turbinate", "turbinate")
condition <- c( "spine", "spine", "spine", "spine", "spine","normal", "normal", "normal","normal", "normal")
metadata <- data.frame(tissue, condition)
rownames(metadata) <- c("SRR5714742","SRR5714765","SRR5714766","SRR5714770","SRR5714774","SRR5714740","SRR5714792","SRR5714793","SRR5714794","SRR5714796")

```

sample names match in both files.
```{r}
all(rownames(metadata) == colnames(data2))
```

DESeq2Dataset object is created. 
```{r}
deseq2Obj <- DESeqDataSetFromMatrix(countData = data2,
                                    colData = metadata,
                                    design = ~ condition)
deseq2Obj <- estimateSizeFactors(deseq2Obj)
sizeFactors(deseq2Obj)

```

normalized counts.
```{r}
#sizeFactors(deseq20bj)
normalized_counts <- counts(deseq2Obj, normalize=TRUE)
View(normalized_counts)

data.frame(normalized_counts)

```

variance stabilizing transformation which transforms counts for data visualization
```{r}
vsd_data <- vst(deseq2Obj, blind = TRUE)
```

hierarchical clustering with correlation heatmaps
```{r}
mat.vsd_data <- assay(vsd_data)
#her bir örnek çifti arasındaki ikili kolerasyon değerleini hesapladım.
cor.vsd_data <- cor(mat.vsd_data)
View(cor.vsd_data)
pheatmap(cor.vsd_data, annotation = dplyr::select(metadata, condition))


```

PCA
```{r}
plotPCA(vsd_data, intgroup="condition")
```


DE analysis
```{r}
deseq2Obj_DE <- DESeq(deseq2Obj)

```

DESeq2 model: mean-variance relation
```{r}
mean_counts <- apply(data2, 1, mean)
variance_counts <-  apply(data2, 1, var)
mean.var.df <- data.frame(mean_counts, variance_counts)
library(ggplot2)
ggplot(mean.var.df) +
  geom_point(aes(x=mean_counts, y=variance_counts)) +
  scale_y_log10() +
  scale_x_log10() +
  xlab("Mean counts per gene") +
  ylab("Variance per gene")
```

dispersion graph
```{r}
plotDispEsts(deseq2Obj_DE)

```


extract results table (alpha threshold is set to 0.05)
```{r}
results(deseq2Obj_DE, alpha = 0.05)

```

groups that we wish to compare are determined to deseq2 and MA plot is geneterated.
```{r}
DE_result <-results(deseq2Obj_DE, 
                    contrast = c("condition", "spine", "normal"),
                    alpha = 0.05)

plotMA(DE_result, ylim=c(-10,10)) #önemli ölçüde ifadesi değişen genler mavi, gri renkler ifadesi değişmeyen genler

```

shrunken LFC 
```{r}
#The apeglm publication demonstrates that 'apeglm' and 'ashr' outperform the original 'normal' shrinkage estimator.
DE_result <- lfcShrink(deseq2Obj_DE,
                       contrast = c("condition", "spine", "normal"),
                       res=DE_result,
                       type = "normal") 
summary(DE_result)
plotMA(DE_result, ylim=c(-10,10)) #önemli ölçüde ifadesi değişen genler mavi, gri renkler ifadesi değişmeyen genler
mcols(DE_result)
head(DE_result, n=10)

```


DE results are arranged.
```{r}
DE_result <-results(deseq2Obj_DE, 
                    contrast = c("condition", "spine", "normal"),
                    alpha = 0.05, 
                    lfcThreshold = 0.58)
summary(DE_result)
```

ensgene varibles are set. 
```{r}
data3 <- data.frame(DE_result)
data3 <- rownames_to_column(data3, var = "ensgene")
data3_res_sig <- subset(data3, padj<0.05)
sig_norm_counts_data <- normalized_counts[data3_res_sig$ensgene,]
```


heatmap
```{r}
heat_colors <- brewer.pal(6, "YlOrRd")
pheatmap(sig_norm_counts_data,
         color = heat_colors,
         cluster_rows = T,
         show_rownames = F,
         annotation = select(metadata, condition),
         scale = "row")
```

volcano plot
```{r}
data3 <- data3 %>%
  mutate(threshold=padj <0.05)

ggplot(data3) +
  geom_point(aes(x=log2FoldChange, y=-log10(padj), color=threshold)) +
  labs(x="log2 fold change", y="-log10 adjusted p-value") +
  ylim(0,15) +
  theme(legend.position = "none",
        plot.title = element_text(size=rel(1.5), hjust = 0.5),
        axis.title = element_text(size=rel(1.25)))
```

expression plots for col2a1 and fn1 genes
```{r}

plotCounts(deseq2Obj, gene="ENSG00000139219")
plotCounts(deseq2Obj
, gene="ENSG00000115414")
```


expression plot for top 20 significant DE genes.
```{r}
top20 <- data.frame(sig_norm_counts_data)[1:20, ] %>%
  rownames_to_column("ensgene")

top20_gathered <- gather(top20, key = "samplename", value = "normalized_counts", 2:7)

top20_gathered_meta <- inner_join(top20_gathered, 
                                  rownames_to_column(metadata, var="samplename"),
                                  by="samplename")

ggplot(top20_gathered_meta) +
  geom_point(aes(x=ensgene, y=normalized_counts, color=condition)) +
  scale_y_log10() +
  xlab("genes") +
  ylab("normalized counts") +
  ggtitle("Top 20 Significant DE Genes") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))

```

